k-plus Anticlustering: An Improved k-means Criterion for Maximizing Between-Group Similarity

### 2023-05-04 paper.Rmd
minor

When defining a new variable $y = (y_1, \ldots, y_N)$ as the squared deviation of each observation from the mean $\overline{\mathbf{x}}$, i.e. $y_i = (x_i - \overline{\mathbf{x}})^2$, the sample variance of $x$ can be reformulated as the mean of $y$:

@@

Research on multiobjective optimization indicates that solutions oftentimes fail to satisfactorily address several criteria if the underlying objectives are in conflict with each other (Ferligoj & Batagelj, 1992). As $\mathit{SSE_{kplus}}$  is a novel criterion, it is so far unknown whether maximizing it can yield a partitioning where both constituting criteria are similar between groups, i.e., whether these objectives are in alignment or conflicting. A motivating example illustrates that aiming to simultaneously optimize similarity of means and variances may be feasible. I created 14 data points following a univariate normal distribution ($M = 15$, $SD = 3$) and then generated all `r n_partitions(14, 2)` possible ways to partition the 14 data points into $K = 2$ equal-sized groups. Figure 1 shows the scatterplot of the criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ across partitions. The partition that is represented by the squared symbol has the overall best highest $\mathit{SSE}$, but is outperformed by many other partitions with regard to $\mathit{SSE_{Var}}$. Similarly, the partition that maximizes $\mathit{SSE_{Var}}$ (represented by the triangular symbol) is not well-suited to optimize $\mathit{SSE}$ at the same time. Thus, when only considering one of the objectives, the mathematically optimal partitioning fails to establish groups that are similar both with regard to their means and variances. The diamond symbol represents the partition that maximizes the unweighted sum of $\mathit{SSE_{Var}}$ and $\mathit{SSE}$ (i.e., $w_1$ = $w_2$ = 1 in the computation of $\mathit{SSE_{kplus}}$), which yields `r df$kmeans[which.max(df$kplus)] / max(df$kmeans) * 100`% of the global maximum $\mathit{SSE}$, and `r df$kvar[which.max(df$kplus)] / max(df$kvar) * 100`% of the global maximum value of $\mathit{SSE_{Var}}$. Thus, even using a very simple unweighted sum approach, the two constituting criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ can be fulfilled to an astonishing degree. This result indicates that the objectives $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ are not in conflict with each other. Moreover, in Figure 1 there is a high concentration of partitions in the upper-right corner where the two objectives are in high agreement.

@@

In both simulation studies, the same local maximum search was applied to optimize the competing anticlustering objectives [Method LCW, @weitz1998]. Building on an initial random allocation, the algorithm proceeds by swapping elements between groups in such a way that each swap improves the objective criterion by the largest possible margin. That is, it starts with the first element and simulates all exchanges with elements that are currently assigned to a different group. After each exchange has been evaluated, the one exchange is realized that improves the objective function the most. No exchange is realized if no improvement is possible. The exchange process is repeated for each element. After that, the procedure restarts at the beginning and is repeated until an iteration through the entire data set no longer yields an improvement. To obtain better results, this local maximum search may be initialized multiple times [@spath1986]. In the simulation studies, five repetitions were used.

@@

In general, the weighting scheme for k-plus anticlustering can become highly complex as it might encompass an arbitrary number of distribution moments, and hence, an arbitrary number of weights. While an exhaustive evaluation of the influence of weights is beyond the scope of the current paper, a small scale simulation study was conducted to investigate the influence of weights for a selected number of moments. The target was the optimization of $\mathit{SSE_{kplusGeneralized}}$ with $T = 4$, i.e. minimizing differences with regard to group means, variances, skewness and kurtosis. Two unweighted approaches were implemented by defining $\mathbf{w} = (1, 1, 1, 1)$. The unweighted standardization approach implemented a $z$-standardization of all columns---the original as well as the augmented variables---of the input data $\mathbf{X'}_{N \times \mathit{TP}}$ before the criterion $\mathit{SSE_{kplusGeneralized}}$ was optimized. In the second case, the unstandardized matrix $\mathbf{X'}_{N \times \mathit{TP}}$ was used. Four weighted approaches were implemented by giving a tenfold increased weight to one of the four moments, respectively. That is, to compute $\mathit{SSE_{kplusGeneralized}}$ favoring similarity in means during the optimization process, I set $\mathbf{w} = (10, 1, 1, 1)$; the weights favoring similarity in variances, skewness and kurtosis, were defined as $\mathbf{w} = (1, 10, 1, 1)$, $\mathbf{w} = (1, 1, 10, 1)$, and $\mathbf{w} = (1, 1, 1, 10)$, respectively. The weighted approaches did not use a standardization of the input data.

@@

## Simulation 2: Comparing k-plus anticlustering to other objectives

In Simulation 2, the performance of two k-plus criteria was compared to the performance of two traditional anticlustering criteria. "Standard" k-plus anticlustering was implemented by maximizing $\mathit{SSE_{kplus}}$, thus minimizing differences with regard to group means and variances. "Generalized" k-plus anticlustering was implemented by maximizing $\mathit{SSE_{kplusGeneralized}}$ using $T = 4$, thus minimizing differences with regard to group means, variances, skewness and kurtosis. To implement the k-plus objectives, the unweighted standardization approach was employed that performed best in Simulation 1. Thus, to optimize a k-plus criterion, the original data set was augmented to incorporate the additional criteria and all variables---the original and the augmented variables---were $z$-standardized. Subsequently, the original k-means criterion $\mathit{SSE}$ was maximized on the basis of the augmented data set. The k-plus methods were compared to (i) k-means anticlustering, minimizing differences with regard to group means, (ii) diversity anticlustering, maximizing the sum of pairwise Euclidean distances within groups, and also (iii) a random allocation of elements to groups.

>>>>

When defining a new variable $y = (y_1, \ldots, y_N)$ as the squared deviation of each observation from the mean $\overline{\mathbf{x}}$, i.e. $y_i = (x_i - \overline{\mathbf{x}})^2$ ($i = 1, \dots, N$), the sample variance of $x$ can be reformulated as the mean of $y$:

@@

Research on multiobjective optimization indicates that solutions oftentimes fail to satisfactorily address several criteria if the underlying objectives are in conflict with each other (Ferligoj & Batagelj, 1992). As $\mathit{SSE_{kplus}}$  is a novel criterion, it is so far unknown whether maximizing it can yield a partitioning where both constituting criteria are similar between groups, i.e., whether these objectives are aligned or in conflict. A motivating example illustrates that aiming to simultaneously optimize similarity of means and variances may be feasible. I created 14 data points following a univariate normal distribution ($M = 15$, $SD = 3$) and then generated all `r n_partitions(14, 2)` possible ways to partition the 14 data points into $K = 2$ equal-sized groups. Figure 1 shows the scatterplot of the criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ across partitions. The partition that is represented by the squared symbol has the overall best highest $\mathit{SSE}$, but is outperformed by many other partitions with regard to $\mathit{SSE_{Var}}$. Similarly, the partition that maximizes $\mathit{SSE_{Var}}$ (represented by the triangular symbol) is not well-suited to optimize $\mathit{SSE}$ at the same time. Thus, when only considering one of the objectives, the mathematically optimal partitioning fails to establish groups that are similar both with regard to their means and variances. The diamond symbol represents the partition that maximizes the unweighted sum of $\mathit{SSE_{Var}}$ and $\mathit{SSE}$ (i.e., $w_1$ = $w_2$ = 1 in the computation of $\mathit{SSE_{kplus}}$), which yields `r df$kmeans[which.max(df$kplus)] / max(df$kmeans) * 100`% of the global maximum $\mathit{SSE}$, and `r df$kvar[which.max(df$kplus)] / max(df$kvar) * 100`% of the global maximum value of $\mathit{SSE_{Var}}$. Thus, even using a very simple unweighted sum approach, the two constituting criteria $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ can be fulfilled to an astonishing degree. This result indicates that the objectives $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ are not in conflict with each other. Moreover, in Figure 1 there is a high concentration of partitions in the upper-right corner where the two objectives are in high agreement.

@@

In both simulation studies, the same local maximum search was applied to optimize the competing anticlustering objectives [Method LCW, @weitz1998]. Building on an initial random allocation, the algorithm proceeds by swapping elements between groups in such a way that each swap improves the objective criterion by the largest possible margin. That is, it starts with the first element and simulates all exchanges with elements that are currently assigned to a different group. After each exchange has been evaluated, the one exchange is realized that improves the objective function the most. No exchange is realized if no improvement is possible. The exchange process is repeated for each element. After that, the procedure restarts at the beginning and is repeated until an iteration through the entire data set no longer yields an improvement. To obtain better results, this local maximum search may be initialized multiple times [@spath1986]. In the simulation studies, five repetitions were used. Note that the local maximum search can be performed on the basis of any arbitrary initial grouping. Specifically, the initial grouping can also be chosen in such a way that the groups are different sized; for example, if the data set cannot evenly be split, it is possible to ensure that the group sizes differ by 1 at most. However, the data sets in the simulation studies were always partitioned into equal-sized groups.

@@

In general, the weighting scheme for k-plus anticlustering can become highly complex as it might encompass an arbitrary number of distribution moments, and hence, an arbitrary number of weights. While an exhaustive evaluation of the influence of weights is beyond the scope of the current paper, a small scale simulation study was conducted to investigate the influence of weights for a selected number of moments. The target was the optimization of $\mathit{SSE_{kplusGeneralized}}$ with $T = 4$, i.e. minimizing differences with regard to group means, variances, skewness and kurtosis. Two unweighted approaches were implemented by defining $\mathbf{w} = (1, 1, 1, 1)$. The unweighted standardization approach implemented a $z$-standardization of all columns---the original as well as the augmented k-plus variables---of the input data $\mathbf{X'}_{N \times \mathit{TP}}$ before the criterion $\mathit{SSE_{kplusGeneralized}}$ was optimized. In the second case, the unstandardized matrix $\mathbf{X'}_{N \times \mathit{TP}}$ was used. Four weighted approaches were implemented by giving a tenfold increased weight to one of the four moments, respectively. That is, to compute $\mathit{SSE_{kplusGeneralized}}$ favoring similarity in means during the optimization process, I set $\mathbf{w} = (10, 1, 1, 1)$; the weights favoring similarity in variances, skewness and kurtosis, were defined as $\mathbf{w} = (1, 10, 1, 1)$, $\mathbf{w} = (1, 1, 10, 1)$, and $\mathbf{w} = (1, 1, 1, 10)$, respectively. The weighted approaches did not use a standardization of the input data.

@@

## Simulation 2: Comparing k-plus to other anticlustering objectives

In Simulation 2, the performance of two k-plus criteria was compared to the performance of two traditional anticlustering criteria. "Standard" k-plus anticlustering was implemented by maximizing $\mathit{SSE_{kplus}}$, thus minimizing differences with regard to group means and variances. "Generalized" k-plus anticlustering was implemented by maximizing $\mathit{SSE_{kplusGeneralized}}$ using $T = 4$, thus minimizing differences with regard to group means, variances, skewness and kurtosis. The unweighted standardization approach, that performed best in Simulation 1, was used as the benchmark implementation for k-plus anticlustering that was compared to the competing objectives. Thus, to optimize a k-plus criterion, the original data set was augmented to incorporate the additional criteria and all variables---the original and the augmented variables---were $z$-standardized. Subsequently, the original k-means criterion $\mathit{SSE}$ was maximized on the basis of the augmented data set. The k-plus methods were compared to (i) k-means anticlustering, minimizing differences with regard to group means, (ii) diversity anticlustering, maximizing the sum of pairwise Euclidean distances within groups, and also (iii) a random allocation of elements to groups.
